\Title

### 2013-08-17 response_to_reviewers.txt

We thank the reviewer for this thoughtful assessment. We have included a
brief discussion of the landscape aspect of HIV fitness and now cite
Ferguson et al, Immunity, 2013.

##############################################################################
# Reviewer #2 (Comments for Author):
##############################################################################

>These are interesting results and the manuscript is clearly written. The second
>observation is not novel, but underappreciated, whereas this is the first
>concrete calculation of the average selective disadvantage of silent mutations
>in HIV taking draft into account. As such, it is necessary for this work to
>establish proper error estimates so that the community can make valid use of
>the results.  With this in mind, the manuscript can be strengthened in the
>following respects:

We thank the reviewer for the constructive comments, which have helped
us to improve the manuscript. We have resorted to a more sophisticated
sampling scheme for parameters in the computer simulations and now scan
a larger parameter space. All our conclusions remain valid, but are based
on more exhaustive sampling. We detail these improvements in our point to point
response below.

@@

##############################################################################
>1) The modeling generation time of 1 day is odd.  It takes about a day for the
>virion to integrate and start transcription.  One usually takes 2 days as the
>mean generation time, but there is individual variation. See next point.

We agree that one generation in HIV takes closer to 2 days than 1
day. Most escape rates and selection coefficients, however, are more
readily measured in days and we have chosen to model HIV evolution with
generation of one day for simplicity. Factor of 2 changes in the
generation have little influence on the dynamics and only renormalize
genetic drift. Genetic drift is not a dominant force in evolution of the
HIV envelope gene. The (effective) population size, the main determinant of drift,
is not known to this accuracy anyway. We have repeated a subset of the
simulations with a generation time of one day, finding no
difference. Our more exhaustive survey of the parameter space includes a
10-fold variation of the population size. Our results are very
insensitive and hence also insensitive to the generation time.

@@

##############################################################################
>2) Related to the above: the simulation does not address how sensitive the
>results are to varying the parameters (in particular, the generation time, the
>outcrossing rate, the mutation rate, and the assigned deleterious fitness of
>the "non-synonymous" changes).  In the absence of described theoretical
>analysis shaping our expectations, I strongly feel that this sensitivity
>analysis needs to be part of the results, and folded into the error bars.

We have greatly extended the parameter space that we survey and now test
different values for the mutation rate, the recombination rate, the
generation time, the fitness effects, and the population size. To
accomodate this high dimensional data, we have changed figure 4. The
figure now shows the joint distribution of the deleterious effect and
the fraction of synonymous mutations that are deleterious that are
compatible with observations as a heat map. The marginal distribution of
the individual parameters consistent with the data are shown in the supplement.

The distribution of allowed deleterious effects is again peaked around
0.002 and the fraction of "synonymous" mutations that can be neutral is
low, as reported in the earlier version of our manuscript.

@@

We agree with reviewer that HIV evolution is bursty, but we would like
to mention that actual number of cells that get infected every day is
around 10^7 and hence much larger than we are able to simulate. The
population size in our simulation can therefore be viewed as an
effective population size. We have investigated the sensitivity of our
results to changes in the population size and find that our conclusions
are insensitive to the population size as long as the product of
population size and mutation rate is not very much smaller than 1.

We have added a discussion of our choice of population size to the
methods section next the part where the other parameter choices are
discussed.

##############################################################################
>6) In the discussions, the rise of synonymous mutations is attributed to draft,
>but the fall to selective effects. In general, some of the fall might also be
>due to overlapping selective sweeps: the manuscript does not discuss this
>possibility.

Yes, much of the stochastic component in the dynamics of the frequencies
of synonymous mutations will be due to draft. However, the consistent
loss of variants, conditional on them being at high frequency earlier,
implies selection against these variants. Without a deleterious effect,
the mutations would fix with a probability equal to their frequency,
regardless of whether they are subject to drift of draft. We have added
a sentence to the discussion to make this point clear.

@@

##############################################################################
>7) The observed neutral fixation pattern of the nonsynonymous mutations is
>mentioned, and two explanations have been proposed to duet of increased
>diversity over synonymous mutations with no increase of fixation over neutral
>expectations.  Neither of the explanations, however, are plausible (at least,
>without argument) absent experimental evidence not mentioned in the text. It
>would perhaps be better to either strengthen this discussion with citation to
>literature, or to weaken it by mentioning these as mere possibilities amongst
>others.

We agree with the reviewer that the both hypothesis we put forward to
explain the fixation pattern of non-synonymous mutations are somewhat
tentative. Qualitatively, however, the processes we discuss have been
observed. We now cite additional experimental evidence (Wei et al,
2003). It is common for escape to happen in multiple equivalent ways and
this has been well documented fro early CTL escape and early nAb escape
(Wei et al, Nature, 2003. Bar et al, PPath, 2012). Data for chronic
infection is less conclusive and selection pressures are generally
smaller, but turnover at targeted epitopes and the replacement of
variants that where previously at high frequencies was observed 3y into
the infection in Wei et al, Nature, 2003.  Whether this fall in
frequency of some non-synonymous variants happens because they are
replaced by a less costly escape variant or because the escape variant
itself is targeted is hard to disentangle. We think it is reasonable
that both of these effects contribute. Whether they are sufficient to
produce the observed fixation patterns of non-synonymous mutations is
largely an empirical question. We simply explored using numerical
experiments whether these effects push the curve in the "right"
direction. We have added a disclaimer that more data is needed to fully
explain the pattern of fixation observed for non-synonymous mutations.

### 2013-08-17 synmut.tex
minor edits

\newpage
\section*{Abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$
T-cells or neutralizing antibodies target a new epitope, the virus often
escapes via nonsynonymous mutations that impair recognition. Synonymous
mutations do not affect this interplay and are often assumed to be
neutral. We test this assumption by tracking synonymous mutations in
longitudinal intrapatient data from the \shankaregion{} part of the
\env{} gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data from SHAPE assays, we find that
synonymous mutations that disrupt base pairs in RNA stems flanking the
variable loops of gp120 are more likely to be lost than other synonymous
changes: these RNA hairpins might be important for HIV-1.  Computational
modeling indicates that, to be consistent with the data, a large fraction
of synonymous mutations in this genomic region needs to be deleterious,
with a selection coefficient of the order of $-0.002$. This weak selection against synonymous
substitutions does not result in a strong pattern of conservation in
cross sectional data, but slows down the rate of evolution
considerably. Our findings are consistent with the notion that large
scale patterns of RNA structure are functionally relevant, while the
precise base pairing pattern is not.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence of
viral proteins in a way that reduces antibody binding or epitope presentation.
Conversely, synonymous mutations do not modify the viral proteins and are
commonly used as approximately neutral markers in studies of viral evolution,
i.e., as a negative control for detecting selected
sites \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes. Several
important RNA secondary structures have been characterized in detail, including the HIV-1
\rev{} response element (RRE) in \env{} which enhances nuclear export of full length
or partially spliced viral transcripts via a complex hairpin RNA structure
\citep{fernandes_hiv-1_2012}. In fact, the HIV-1 genome is full of RNA
structures \citep{watts_architecture_2009} with no or unknown
function. However, large scale modification of secondary structures
can results in substantial reduction of the replication capacity
\citep{keating_rich_2009} and the propensity of forming RNA stems
anticorrelates with the rate of evolution
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}. 
These poorly characterized RNA structures are conserved to varying degree between HIV-1 and
SIV: corresponding regions tend to be part
of similar structural elements, but individual base pairings are very
rarely conserved \citep{pollom_comparison_2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimates of time to fixation/extinction, single nucleotide variants
(SNVs) were binned by
frequency and the time to first reaching either fixation or extinction was
stored. The fixation probability was determined as the long-time limit of the
resulting curves. Mutations that reached high frequency but neither fixed nor
were lost were classified as ``floating'', with one exception: if they first
reached high frequencies within 3 years of the last time point, it was assumed
they had not had sufficient time to settle, so they were discarded.

@@

\subsubsection*{Parameters}
In order to simulate HIV, we have to specify several parameters that are
known to various degrees. The mutation rate has been measured using
exogenous LacZ constructs and the average nucleotide substitution rate is 
estimated to be $\mu\approx 2 \times 10^{-5}$ per generation \citep{mansky_lower_1995,
abram_nature_2010}. In our simulations, we vary the mutations rate
from $10^{-5} - 3\times 10^{-5}$ per day. Different virions within an
infected person can recombine their genome by coinfection of the same
cell, followed by template switching during reverse transcription. 
Template switching happens between around ten times per reverse
transcription \citep{levy_dynamics_2004}. The combined recombination rate of template
switching and coinfection has been estimated based on modeling to be around $\rho\approx
10^{-5}$ per nucleotide and day
\citep{neher_recombination_2010,batorsky_estimate_2011}, implying a
coinfection rate of a few percent or less, as has been recently
confirmed experimentally \citep{josefsson_single_2013}. We assume a
template switching rate of $10^{-3}$ per nucleotide and vary the
recombination rate between $7\times 10^{-6}- 3\times 10^{-5} 10^{-5}$ by adjusting the
coinfection rate. The population size relevant to evolution in chronic
infection has been hotly debated. The number of cells infected by HIV-1
per day is on the order of $10^{7}$ \citep{perelson_hiv-1_1996}, but the
number of viruses contributing to the next generation might be
considerably smaller due to the burstiness of replication. 
Recent evidence points to a relevant population
size in excess of the inverse mutation rate
\citep{boltz_ultrasensitive_2012}. In fact, the evolutionary dynamics
depends only weakly on the actual value of the population size once the
product $N\mu$ is of order one or larger \citep{neher_genetic_2011}. We
simulate populations between $N=10^4$ and $10^5$. Larger populations
become prohibitively costly to simulate.

Another crucial ingredient for our simulations is the fitness landscape,
i.e., the effects of mutations on fitness and possible interactions
between mutations. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $\alpha$ or a deleterious effect $s_d$ with probability
$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 1.
Mutations at the first and second positions were assigned strongly deleterious 
fitness effects of $-0.01$ (simulations with a larger cost of $-0.1$ were also
performed and produced similar results). At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. The rate of escape in chronic infection is
reported to be on the order of a few percent
\citep{Asquith:2006p28003,moore_limited_2009}, consistent with the
finding that virus is neutralized by serum from 3-6 month later
\citep{richman_rapid_2003}. We simulated escape rates between $2 \cdot
10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
The rate at which new antibody challenges arise is more difficult to
quantify, but a substantial fraction of nonsynonymous substitutions in gp120 are
probably driven by escape. Since the sequence of the C2-V5 region of
gp120  changes at a rate of about 1\% per year
\citep{shankarappa_consistent_1999}, this rate should be on the order of
$k_A=0.005$ per day. We simulate a range from $10^{-4}$ to $10^{-2}$ per day.

@@

\subsubsection*{Sampling and analysis}
To obtain reasonable sampling of a particular parameter combination we
ran simulation for 6000 days and we repeated each run 100 times with
different seeds for the random number generator. Both full-length HIV-1
genomes and \env{}-only simulations were performed and yielded
comparable results. Populations were initialized with a homogeneous
founder population and were kept at an average size of $N=10^4$
throughout the simulation. After 30 generations of burn-in to create
genetic diversity, new epitopes were introduced at later times according 
to an exponential distribution with rate
$k_A$. The simulations were repeated 3000 times with a seven dimensional
Latin hypercube sampling scheme \citep{mckay_comparison_1979}, bounded by the ranges
given for each of these parameters above.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We studied the dynamics and fate of synonymous variants more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citet{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were not considered; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculated the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in panel \ref{fig:fixp2}. The neutral result is
shown as a black dashed line. We found that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion{} (red line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we found no such reduction in $\pfix$ (green line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}
and show that synonymous mutations fix less often then expected under neutrality. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
line) -- a point to which we come back below.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for a reduced fixation of synonymous variants in
\shankaregion{} is secondary structures in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious SNVs reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation of some deleterious synonymous variants is not fully
unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
and green lines in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,wei_antibody_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
cohort.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
SNVs at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $-0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)

@@

\FIG{simfixpvar} shows simulation results for the fixation probability and the
synonymous diversity for different deleterious effects sizes of synonymous mutations.
We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
with an synonymous SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure. To quantify the depression
of the fixation probability, we calculate the area between the measured fixation
probability and the diagonal, which is the neutral expectation
(\FIG{simfixpvar}, lower inset). If no fixation happens, the area will be
$-0.5$; if every SNV fixes, the area will be $+0.5$. In HIV-1 infected
patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.2$
for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
probability of synonymous SNVs decreases from the neutral expectation
($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.5$) as the
fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
deleterious SNVs are selected against.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution for many different combinations of
effects of synonymous mutations, $s_d$, and the fraction of synonymous
sites that are neutral, $\alpha$, escape rates and number of targeted epitopes. In
addition, we vary the mutation rate, recombination rate and population
size (see methods).  Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Figure \ref{fig:simsfig} shows
the distribution of deleterious effects $s_d$ and neutral fraction
$\alpha$ for which we found $A_\text{syn} < -0.15$ and $0.0025 <
P_\text{interm} < 0.010$. This subset of parameters indicates that in
order to be consistent with the data only
a minority of synonyous sites ($\lesssim 0.2$) can be neutral  and that
the deleterious effects of the remainder is on the order of $|s_d| \sim
0.002$.

>>>>

\newpage
\section*{Abstract}
\noindent
Intrapatient HIV-1 evolution is driven by the adaptive immune system
resulting in rapid change of HIV-1 proteins. When cytotoxic CD8${}^+$
T-cells or neutralizing antibodies target a new epitope, the virus often
escapes via nonsynonymous mutations that impair recognition. Synonymous
mutations do not affect this interplay and are often assumed to be
neutral. We test this assumption by tracking synonymous mutations in
longitudinal intrapatient data from the \shankaregion{} part of the
\env{} gene. We find that most synonymous variants are lost even though
they often reach high frequencies in the viral population, suggesting a
cost to the virus. Using published data from SHAPE assays, we find that
synonymous mutations that disrupt base pairs in RNA stems flanking the
variable loops of gp120 are more likely to be lost than other synonymous
changes: these RNA hairpins might be important for HIV-1.  Computational
modeling indicates that, to be consistent with the data, a large fraction
of synonymous mutations in this genomic region needs to be deleterious,
with a cost of the order of $0.002$ per day. This weak selection against synonymous
substitutions does not result in a strong pattern of conservation in
cross sectional data, but slows down the rate of evolution
considerably. Our findings are consistent with the notion that large
scale patterns of RNA structure are functionally relevant, while the
precise base pairing pattern is not.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
HIV-1 evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via cytotoxic CD8${}^+$ T-cells (CTLs) and neutralizing antibodies
(nAbs)~\citep{rambaut_causes_2004} and facilitated by HIV-1's high mutation rate
~\citep{mansky_lower_1995,abram_nature_2010}. Escape mutations
in epitopes targeted by CTLs are typically observed during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving parts of the HIV-1 genome are the variable
loops V1-V5 in the envelope protein gp120 (V loops), which change to avoid recognition by
nAbs. Escape mutations in \env{}, the gene encoding gp120, spread through the
viral population within a few months.
Consistent with this time scale, it is found that serum from a particular time
typically neutralizes autologous virus extracted more than 3-6 months earlier but not contemporary
virus \citep{richman_rapid_2003}.

Escape mutations are selected because they change the amino acid sequence of
viral proteins in a way that reduces antibody binding or epitope presentation.
Conversely, synonymous mutations do not modify the viral proteins and are
commonly used as approximately neutral markers in studies of viral evolution,
i.e., as a negative control for detecting selected
sites \citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}. In addition to
maintaining protein function and avoiding the adaptive immune recognition,
however, the HIV-1 genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate
at the RNA level and are sensitive to synonymous changes. Several
important RNA secondary structures have been characterized in detail, including the HIV-1
\rev{} response element (RRE) in \env{} which enhances nuclear export of full length
or partially spliced viral transcripts via a complex hairpin RNA structure
\citep{fernandes_hiv-1_2012}. In fact, the HIV-1 genome is full of RNA
structures \citep{watts_architecture_2009} with no or unknown
function. However, large scale modification of secondary structures
can result in substantial reduction of the replication capacity
\citep{keating_rich_2009} and the propensity of forming RNA stems
anticorrelates with the rate of evolution
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}. 
These poorly characterized RNA structures are conserved to varying degree between HIV-1 and
SIV: corresponding regions tend to be part
of similar structural elements, but individual base pairings are very
rarely conserved \citep{pollom_comparison_2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Fixation probability and secondary structure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the estimates of time to fixation/extinction, single nucleotide variants
(SNVs) were binned by
frequency and the time to first reaching either fixation or extinction was
stored. The fixation probability was determined as the long-time limit of the
resulting curves. Mutations that reached high frequency but neither fixed nor
were lost were classified as ``floating'' unless they first
reached high frequencies within 3 years of the last time point. In that case, it was assumed
they had not had sufficient time to fix and were discarded.

@@

\subsubsection*{Parameters}
In order to simulate HIV, we have to specify several parameters that are
known to various degrees. The mutation rate has been measured using
exogenous LacZ constructs and the average nucleotide substitution rate is 
estimated to be $\mu\approx 2 \times 10^{-5}$ per generation \citep{mansky_lower_1995,
abram_nature_2010}. In our simulations, we vary the mutations rate
from $10^{-5} - 3\times 10^{-5}$ per day. Different virions within an
infected person can recombine their genome by coinfection of the same
cell, followed by template switching during reverse transcription. 
Template switching happens between around ten times per reverse
transcription \citep{levy_dynamics_2004}. The combined recombination rate of template
switching and coinfection has been estimated based on modeling to be around $\rho\approx
10^{-5}$ per nucleotide and day
\citep{neher_recombination_2010,batorsky_estimate_2011}, implying a
coinfection rate of a few percent or less, as has been recently
confirmed experimentally \citep{josefsson_single_2013}. We assume a
template switching rate of $10^{-3}$ per nucleotide and vary the
recombination rate between $5\times 10^{-6}- 5\times 10^{-5}$ by adjusting the
coinfection rate. The population size relevant to evolution in chronic
infection has been hotly debated. The number of cells infected by HIV-1
per day is on the order of $10^{7}$ \citep{perelson_hiv-1_1996}, but the
number of viruses contributing to the next generation might be
considerably smaller due to the burstiness of replication. 
Recent evidence points to a relevant population
size in excess of the inverse mutation rate
\citep{boltz_ultrasensitive_2012}. In fact, the evolutionary dynamics
depends only weakly on the actual value of the population size once the
product $N\mu$ is of order one or larger \citep{neher_genetic_2011}. We
simulate populations between $N=10^4$ and $10^5$. Larger populations
become prohibitively costly to simulate.

Another crucial ingredient for our simulations is the fitness landscape,
i.e., the effects of mutations on fitness and possible interactions
between mutations. For simplicity, third positions of every codon were
deemed synonymous and assigned either a selection coefficient $0$ with
probability $\alpha$ or a deleterious effect $s_d$ with probability
$1-\alpha$. In our simulations, $\alpha$ varies from 0 to 1.
Mutations at the first and second positions were assigned deleterious 
fitness effects of $-0.01$ (simulations with a larger cost of $-0.1$ were also
performed and produced similar results). At 
rate $k_A$, a random locus in the genome is designated an epitope that can
escape by one or several mutations with exponentially distributed escape rates
with mean $\epsilon$. The rate of escape in chronic infection is
reported to be on the order of a few percent
\citep{Asquith:2006p28003,moore_limited_2009}, consistent with the
finding that virus is neutralized by serum from 3-6 month later
\citep{richman_rapid_2003}. We simulated escape rates between $2 \cdot
10^{-2.5}$ and $2 \cdot 10^{-1.5}$ per day. 
The rate at which new antibody challenges arise is more difficult to
quantify, but a substantial fraction of nonsynonymous substitutions in gp120 are
probably driven by escape. Since the sequence of the C2-V5 region of
gp120  changes at a rate of about 1\% per year
\citep{shankarappa_consistent_1999}, this rate should be on the order of
$k_A=0.005$ per day. We simulate a range from $10^{-4}$ to $2\times 10^{-2}$ per day.

@@

\subsubsection*{Sampling and analysis}
To obtain reasonable sampling of a particular parameter combination we
ran simulation for 6000 days and we repeated each run 100 times with
different seeds for the random number generator. Both full-length HIV-1
genomes and \env{}-only simulations were performed and yielded
comparable results. Populations were initialized with a homogeneous
founder population. After 30 generations of burn-in to create
genetic diversity, new epitopes were introduced at random with rate
$k_A$. The simulations were repeated 3000 times with a seven dimensional
Latin hypercube sampling scheme \citep{mckay_comparison_1979}, bounded by the ranges
given for each of these parameters above. For all parameter but the
fraction $\alpha$ of neutral sites,  parameters were sampled uniformly in
log-space.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Many synonymous SNVs in \shankaregion{} are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We studied the dynamics and fate of synonymous variants more quantitatively by
analyzing data from seven patients from \citet{shankarappa_consistent_1999} and
\citet{liu_selection_2006} as well as three patients from
\citet{bunnik_autologous_2008} (patients with strong viral population structure
were not considered; see methods and \figurename~S\PCApat). The
former data set is restricted to the \shankaregion{} region of \env, while the
data from \citet{bunnik_autologous_2008} cover most of \env.  We considered all
SNVs in a frequency interval $[\nu_0-\delta\nu, \nu_0+\delta\nu]$ at some time
$t$, and calculated the fraction that are still observed at later times $t+\Delta
t$. Plotting this fraction against the time interval $\Delta t$, we see that
most synonymous SNVs segregate for roughly one year and are lost much more
frequently than expected under neutrality (panel \ref{fig:fixp1}). The long-time
probability of fixation, $\pfix$, is shown as a function of the initial
frequency $\nu_0$ in panel \ref{fig:fixp2}. The neutral expectation is
shown as a black dashed line. We found that $\pfix$ of synonymous
variants is far below the neutral expectation in \shankaregion{} (red line).
Outside of \shankaregion, using data from \citet{bunnik_autologous_2008} only,
we found no such reduction in $\pfix$ (green line). Restricted to the
\shankaregion{} region, the sequence samples from \citet{bunnik_autologous_2008}
are fully compatible with data from \citet{shankarappa_consistent_1999}
and show that synonymous mutations fix less often then expected under neutrality. The
nonsynonymous SNVs seem to follow more or less the neutral expectation (blue
line) -- a point to which we come back below.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Synonymous mutations in \shankaregion{} tend to disrupt RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for a reduced fixation of synonymous variants in
\shankaregion{} is secondary structure in the viral RNA, the disruption of which
is deleterious to the virus \citep{forsdyke_reciprocal_1995,
snoeck_mapping_2011, sanjuan_interplay_2011}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Deleterious SNVs can reach high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation of some deleterious synonymous variants is not fully
unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts of the
genome (in \shankaregion{} but not in the rest of \env{}; compare the red
and green lines in \FIG{fixp2}).
The region \shankaregion{} undergoes frequent adaptive changes to evade
recognition by neutralizing antibodies \cite{williamson_adaptation_2003,wei_antibody_2003,
richman_rapid_2003}. Due to the limited amount of recombination in HIV-1
\cite{neher_recombination_2010, batorsky_estimate_2011}, deleterious variants
that are linked to adaptive variants can reach high frequency. This process is
known as hitchhiking \citep{smith_hitch-hiking_1974} or genetic draft
\citep{gillespie_genetic_2000,neher_genetic_2011}. Hitchhiking is apparent in
\FIG{aft}, which shows that many SNVs change rapidly in frequency as a
cohort.

The approximate magnitude of the deleterious effects can be estimated from
\FIG{fixp1}, which shows the distribution of times after which synonymous
SNVs at intermediate frequencies become fixed or lost. The typical time to
loss is of the order of 500 days. If this loss is driven by the deleterious
effect of the mutation, this corresponds to deleterious effects $s_d$ of the
order of $0.002$ per day. (This is only an average estimate: every single
mutation is expected to have a slightly different fitness effect.)

@@

\FIG{simfixpvar} shows simulation results for the fixation probability and the
synonymous diversity for different deleterious effects sizes of synonymous mutations.
We quantify synonymous diversity via $P_\text{interm}$, the fraction of sites
with an synonymous SNV at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure. To quantify the depression
of the fixation probability, we calculate the area between the measured fixation
probability and the diagonal, which is the neutral expectation
(\FIG{simfixpvar}, lower inset). We restrict this area up to frequency
$\nu=0.5$ since the high frequency part is rather noisy.
If no fixation happens, this restricted area will be
$0.125$; if every SNV fixes, the area will be $0.125$. In HIV-1 infected
patients, we find $P_\text{interm} \approx 0.01$, $A_\text{syn} \approx -0.08$
for synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous
changes. In the three simulations shown in \FIG{simfixpvar}, the fixation
probability of synonymous SNVs decreases from the neutral expectation
($A_\text{syn} \approx 0$) to zero ($A_\text{syn} \approx -0.125$) as the
fitness cost of the SNVs increases; the synonymous diversity plummets as well, as
deleterious SNVs are selected against.

To map the parameter range of the model that is compatible with the data, we
repeatedly simulated the evolution for many different combinations of
effects of synonymous mutations, $s_d$, and the fraction of synonymous
sites that are neutral, $\alpha$, escape rates and number of targeted epitopes. In
addition, we vary the mutation rate, recombination rate and population
size (see methods).  Among all simulations, we selected the ones
that show $A_\text{syn}$ and $P_\text{interm}$ as observed in the data, i.e., a
large depression in fixation probability of synonymous SNVs but, simultaneously,
a moderately high synonymous diversity. Figure \ref{fig:simsfig} shows
the distribution of deleterious effects $s_d$ and neutral fraction
$\alpha$ for which we found $A_\text{syn} < -0.15$ and $0.005 <
P_\text{interm} < 0.015$. This subset of parameters indicates that in
order to be consistent with the data only
a minority of synonymous sites ($\alpha \lesssim 0.2$) can be neutral  and that
the deleterious effects of the remainder is on the order of $s_d \sim
0.002$.
